---
title: "02_traditional_models"
format: html
editor: visual
---

```{r setup}
#| message: false
#| warning: false

# ── Standalone setup ──────────────────────────────────────────────────────────
# Split out of the monolithic thesis notebook, so it loads its own libraries and
# rebuilds the modelling frames from the shared table that 01_data_prep.qmd
# writes. Paths are relative to this file (analysis/): the table lives in
# ../artifacts, results are written to ../results.
pkgs <- c("tidyverse", "quanteda", "quanteda.textmodels", "caret", "e1071", "randomForest")
missing <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(missing) > 0) install.packages(missing)
invisible(lapply(pkgs, library, character.only = TRUE))

art_dir <- "../artifacts"
res_dir <- "../results"
tab_dir <- "../results/tables"
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)

# Shared modelling table (built by 01_data_prep.qmd; requires the dataset).
# Classical ML trains on the full 80% (train + validation) and scores the 20% test.
model_data <- readr::read_csv(file.path(art_dir, "gossip_model_data.csv"),
                              show_col_types = FALSE) %>%
  mutate(
    subclass = factor(subclass, levels = c("HF", "MF", "HR", "MR")),
    label    = factor(label,    levels = c("fake", "real")),
    split    = factor(split,    levels = c("train", "validation", "test"))
  )
trainml_df <- model_data %>% filter(split %in% c("train", "validation"))   # 80%
test_df    <- model_data %>% filter(split == "test")                       # 20%
cat(sprintf("Loaded model_data: trainml=%d  test=%d\n", nrow(trainml_df), nrow(test_df)))
```

# TF-IDF

## Build document text and corpora

```{r tfidf-corpora}
# ── Build corpora from the shared modelling frames ───────────
# trainml_df (80%) and test_df (20%) come from the data-prep split and already carry doc_text.
# Classical models have nothing to checkpoint, so they train on the full 80%.

train_src <- trainml_df %>% select(doc_id, label, subclass, doc_text)
test_src  <- test_df    %>% select(doc_id, label, subclass, doc_text)

stopifnot(!any(is.na(train_src$doc_text)), !any(train_src$doc_text == ""))
stopifnot(!any(is.na(test_src$doc_text)),  !any(test_src$doc_text  == ""))

train_corpus <- quanteda::corpus(train_src, text_field = "doc_text", docid_field = "doc_id")
test_corpus  <- quanteda::corpus(test_src,  text_field = "doc_text", docid_field = "doc_id")

cat("Train corpus:", quanteda::ndoc(train_corpus), "\n")
cat("Test corpus: ", quanteda::ndoc(test_corpus),  "\n")
```

## Tokenisation function

```{r tfidf-tokenise}
# ── Tokenisation ─────────────────────────────────────────────

# Start with no custom stopwords.
# After checking topfeatures(), we can add obvious boilerplate words here.
custom_stop <- character(0)

make_tokens <- function(corp) {
  corp %>%
    quanteda::tokens(
      remove_punct   = TRUE,
      remove_numbers = TRUE,
      remove_symbols = TRUE,
      remove_url     = TRUE
    ) %>%
    quanteda::tokens_tolower() %>%
    quanteda::tokens_remove(quanteda::stopwords("en")) %>%
    quanteda::tokens_remove(custom_stop) %>%
    quanteda::tokens_wordstem(language = "english")
}

train_toks <- make_tokens(train_corpus)
test_toks  <- make_tokens(test_corpus)

cat("Tokenisation done.\n")
```

## Fit vocabulary and TF-IDF on train only

```{r tfidf-vocab}
# ── Train-fitted vocabulary + leakage-free TF-IDF ────────────

# 1) Build raw TRAIN document-feature matrix
train_dfm_raw <- quanteda::dfm(train_toks)

cat("Raw train vocabulary:", quanteda::nfeat(train_dfm_raw), "features\n")

# 2) Trim vocabulary using TRAIN only
# min_docfreq = 5 removes extremely rare words.
# max_docfreq = 0.50 removes very common terms appearing in more than 50% of documents.
train_dfm_trimmed <- train_dfm_raw %>%
  quanteda::dfm_trim(min_docfreq = 5, docfreq_type = "count") %>%
  quanteda::dfm_trim(max_docfreq = 0.50, docfreq_type = "prop")

cat("Vocabulary after docfreq trimming:",
    quanteda::nfeat(train_dfm_trimmed), "features\n")

# 3) Cap dimensionality so glm / naiveBayes / randomForest remain manageable
max_features <- 1000

if (quanteda::nfeat(train_dfm_trimmed) > max_features) {
  keep_features <- names(quanteda::topfeatures(train_dfm_trimmed, n = max_features))
  train_dfm <- quanteda::dfm_select(
    train_dfm_trimmed,
    pattern   = keep_features,
    selection = "keep",
    valuetype = "fixed"
  )
} else {
  train_dfm <- train_dfm_trimmed
}

cat("Final vocabulary:", quanteda::nfeat(train_dfm), "features\n")

# Inspect most frequent terms in the final training vocabulary.
# If you see obvious boilerplate, add it to custom_stop and re-run the tokenisation.
print(quanteda::topfeatures(train_dfm, 30))

# 4) Compute IDF on TRAIN only
train_idf <- quanteda::docfreq(
  train_dfm,
  scheme = "inverse",
  base   = 10
)

# 5) TRAIN TF-IDF = train counts × train IDF
train_tfidf <- quanteda::dfm_weight(
  train_dfm,
  weights = train_idf
)

# 6) TEST counts: same tokenisation, but force TRAIN vocabulary
test_dfm_raw <- quanteda::dfm(test_toks)

test_dfm <- quanteda::dfm_match(
  test_dfm_raw,
  features = quanteda::featnames(train_dfm)
)

# 7) TEST TF-IDF = test counts × TRAIN IDF
test_tfidf <- quanteda::dfm_weight(
  test_dfm,
  weights = train_idf
)

cat("TF-IDF matrices built.\n")
```

## Convert to model-ready objects

```{r tfidf-model-ready}
# ── Convert to model-ready data ──────────────────────────────

# Labels and subclasses stay aligned with the dfm rows through docvars.
# Level order is c("real", "fake") — i.e. "fake" is the SECOND level — set once
# here so every model downstream treats "fake" as the positive class. In
# particular glm() models P(second level), so this makes glm predict P(fake)
# with no per-model releveling. (Naive Bayes and the random forest pick the
# fake-probability column by name, so they are insensitive to the order anyway.)
train_y <- factor(
  quanteda::docvars(train_tfidf, "label"),
  levels = c("real", "fake")
)

test_y <- factor(
  quanteda::docvars(test_tfidf, "label"),
  levels = c("real", "fake")
)

train_sub <- factor(
  quanteda::docvars(train_tfidf, "subclass"),
  levels = c("HF", "MF", "HR", "MR")
)

test_sub <- factor(
  quanteda::docvars(test_tfidf, "subclass"),
  levels = c("HF", "MF", "HR", "MR")
)

# Dense matrices for glm, e1071::naiveBayes, and randomForest
train_x <- quanteda::convert(train_tfidf, to = "matrix")
test_x  <- quanteda::convert(test_tfidf,  to = "matrix")

# Raw-count matrices over the SAME train-fitted vocabulary.
# The proposal promised TF-IDF *and* count-vectorised features; counts are also
# the natural input for Naive Bayes. train_dfm/test_dfm already hold the trimmed,
# train-matched counts (before TF-IDF weighting), so these share identical
# columns and row order with the TF-IDF matrices above.
train_counts_x <- quanteda::convert(train_dfm, to = "matrix")
test_counts_x  <- quanteda::convert(test_dfm,  to = "matrix")

# Make feature names syntactically valid for formula-based models
clean_feature_names <- make.names(colnames(train_x), unique = TRUE)

colnames(train_x) <- clean_feature_names
colnames(test_x)  <- clean_feature_names
colnames(train_counts_x) <- clean_feature_names
colnames(test_counts_x)  <- clean_feature_names

# Model-ready data frames (TF-IDF features)
train_model_df <- as.data.frame(train_x) %>%
  mutate(label = train_y)

test_model_df <- as.data.frame(test_x) %>%
  mutate(label = test_y)

# Model-ready data frames (count features)
train_counts_df <- as.data.frame(train_counts_x) %>%
  mutate(label = train_y)

test_counts_df <- as.data.frame(test_counts_x) %>%
  mutate(label = test_y)
```

## Sanity checks

```{r tfidf-sanity}
# ── Sanity checks ────────────────────────────────────────────

stopifnot(identical(colnames(train_x), colnames(test_x)))
stopifnot(identical(colnames(train_counts_x), colnames(test_counts_x)))
stopifnot(identical(colnames(train_x), colnames(train_counts_x)))

stopifnot(nrow(train_x) == length(train_y))
stopifnot(nrow(test_x)  == length(test_y))

stopifnot(nrow(train_x) == nrow(train_model_df))
stopifnot(nrow(test_x)  == nrow(test_model_df))

stopifnot(!any(is.na(train_x)))
stopifnot(!any(is.na(test_x)))
stopifnot(!any(is.na(train_counts_x)))
stopifnot(!any(is.na(test_counts_x)))

cat("train_x:", nrow(train_x), "x", ncol(train_x), "\n")
cat("test_x: ", nrow(test_x),  "x", ncol(test_x), "\n")

cat("\nTrain label balance:\n")
print(prop.table(table(train_y)))

cat("\nTest label balance:\n")
print(prop.table(table(test_y)))

cat("\nTest subclass counts:\n")
print(table(test_sub))

cat("\nFeature mismatch test -> train:",
    length(setdiff(colnames(test_x), colnames(train_x))), "\n")
```

The sections below train the classical models and write one frozen prediction file per model to ../artifacts. The "Evaluation" section at the end scores them from those files (no retraining) — exactly as 03 scores BERT and 04 the LLMs from their own frozen predictions.

# Logistic regression

## train

```{r my-logistic_training}
# ── Logistic regression ─────────────────────────────────────
# label is already levelled real < fake, so glm predicts P(fake).

set.seed(2026)

logit_model <- glm(
  label ~ .,
  data    = train_model_df,
  family  = binomial(link = "logit"),
  control = glm.control(maxit = 100)
)

summary(logit_model)
```

## test

```{r logit-test}
# ── Test predictions ────────────────────────────────────────

logit_prob_fake <- predict(
  logit_model,
  newdata = test_model_df,
  type = "response"
)

summary(logit_prob_fake)

logit_pred <- factor(
  ifelse(logit_prob_fake >= 0.5, "fake", "real"),
  levels = c("real", "fake")
)
```

# Naive Bayes

We fit Naive Bayes in two forms, each paired with the representation it is built for. The first is a **Gaussian** NB (`e1071::naiveBayes`) on the TF-IDF matrix, kept for a like-for-like comparison with the other TF-IDF models. The second is a **multinomial** NB (`quanteda.textmodels::textmodel_nb`) on the raw counts — the classical text-classification NB and the model the proposal's count features were intended for. The two representations are not interchangeable here. Gaussian NB is invariant to the per-term IDF rescaling: multiplying a feature by a positive constant rescales its fitted mean and standard deviation by the same constant, so the per-class density changes by a factor common to both classes and the decision is unchanged. Fitting Gaussian NB on counts would therefore reproduce the TF-IDF run exactly. Multinomial NB, by contrast, models term counts directly and does respond to the representation, so pairing each representation with its proper classifier is what makes the comparison informative rather than a tautology.

## TF-IDF features

```{r nb-tfidf}
# ── Naive Bayes on TF-IDF features ──────────────────────────

set.seed(2026)

nb_model <- e1071::naiveBayes(label ~ ., data = train_model_df)

nb_pred <- factor(
  predict(nb_model, newdata = test_model_df, type = "class"),
  levels = c("real", "fake")
)

nb_prob_fake <- predict(nb_model, newdata = test_model_df, type = "raw")[, "fake"]

summary(nb_prob_fake)
```

## Count features

```{r nb-counts}
# ── Multinomial Naive Bayes on raw-count features ────────────
# textmodel_nb is the classical text NB. It consumes the count dfm directly:
# train_dfm / test_dfm are the trimmed, train-matched COUNT dfms,
# before any TF-IDF weighting. train_y is levelled c("real","fake") and is row-
# aligned with train_dfm. Class priors are irrelevant here (training is 50/50),
# so the default uniform prior and Laplace smoothing (smooth = 1) are kept.
set.seed(2026)

nb_counts_model <- quanteda.textmodels::textmodel_nb(
  x            = train_dfm,
  y            = train_y,
  distribution = "multinomial"
)

nb_counts_pred <- factor(
  predict(nb_counts_model, newdata = test_dfm, type = "class"),
  levels = c("real", "fake")
)

nb_counts_prob_fake <- predict(
  nb_counts_model, newdata = test_dfm, type = "probability"
)[, "fake"]

summary(nb_counts_prob_fake)
```

The Gaussian NB on TF-IDF tends to over-predict the fake class, achieving high recall at the cost of poor specificity (many false positives among both human- and machine-written real news). The multinomial NB on counts is the standard text formulation and should be read as the fair Naive Bayes baseline; the gap between the two isolates how much of the Gaussian variant's weakness is an artifact of fitting a normal density to sparse, zero-inflated count features rather than a limitation of Naive Bayes as a family.

<!-- Naive Bayes metrics are computed in the Evaluation section below, from the frozen preds. -->


# Random Forest

```{r rf-train}
# ── Random Forest ───────────────────────────────────────────
# train_y / test_y are already factors levelled real < fake.

p_rf <- ncol(train_x)

set.seed(2026)

rf_model <- randomForest::randomForest(
  x          = train_x,
  y          = train_y,
  ntree      = 500,
  mtry       = floor(sqrt(p_rf)),
  importance = TRUE,
  do.trace   = 50
)

rf_model
```

## Predict on the test set

```{r rf-test}
# ── Test predictions ────────────────────────────────────────

rf_pred <- predict(
  rf_model,
  newdata = test_x,
  type = "class"
)

rf_pred <- factor(rf_pred, levels = c("real", "fake"))

rf_prob <- predict(
  rf_model,
  newdata = test_x,
  type = "prob"
)

rf_prob_fake <- rf_prob[, "fake"]

summary(rf_prob_fake)
```

## OOB error curve

```{r rf-oob}
# ── OOB error curve data (figure rendered in the Evaluation section) ─
rf_oob <- tibble(
  trees = seq_len(nrow(rf_model$err.rate)),
  OOB   = rf_model$err.rate[, "OOB"],
  real  = rf_model$err.rate[, "real"],
  fake  = rf_model$err.rate[, "fake"]
)
readr::write_csv(rf_oob, file.path(art_dir, "rf_oob.csv"))
```

## Variable importance

```{r rf-importance}
# ── RF variable importance (permutation, MeanDecreaseAccuracy) ──────
# Model internal — not derivable from test predictions — so write it here for
# the Evaluation section below to render the top-20 figure.
rf_importance <- randomForest::importance(rf_model) %>%
  as.data.frame() %>%
  rownames_to_column("Feature") %>%
  arrange(desc(MeanDecreaseAccuracy))
readr::write_csv(rf_importance, file.path(art_dir, "rf_importance.csv"))
```

## Frozen per-row predictions

```{r ml-freeze-preds}
# ── Frozen per-row test predictions (one file per model) ────────────
# Same shape as bert_test_predictions.csv and llm_preds_*.csv so every model
# family follows one pipeline: doc_id + subclass + truth + pred + prob_fake, with
# NO article text. Committed to ../artifacts, these let the evaluation/comparison
# be reproduced without retraining (the parallel of the BERT/LLM frozen preds).
ml_test_doc_id <- quanteda::docnames(test_tfidf)   # aligned with every pred vector below

ml_pred_specs <- list(
  logit     = list(pred = logit_pred,     prob = logit_prob_fake),
  nb_tfidf  = list(pred = nb_pred,        prob = nb_prob_fake),
  nb_counts = list(pred = nb_counts_pred, prob = nb_counts_prob_fake),
  rf        = list(pred = rf_pred,        prob = rf_prob_fake)
)

for (key in names(ml_pred_specs)) {
  tibble(
    doc_id    = ml_test_doc_id,
    subclass  = as.character(test_sub),
    truth     = as.character(test_y),
    pred      = as.character(ml_pred_specs[[key]]$pred),
    prob_fake = as.numeric(ml_pred_specs[[key]]$prob)
  ) %>% readr::write_csv(file.path(art_dir, sprintf("ml_preds_%s.csv", key)))
}
message("✓ Wrote frozen ML predictions -> ", art_dir,
        "/ml_preds_{logit,nb_tfidf,nb_counts,rf}.csv")
```

# Evaluation

Each classical model is scored from its committed `ml_preds_*.csv` — no retraining
— so this section reproduces for free and uses the same helpers and reading pattern
as the BERT (`03`) and LLM (`04`) evaluations.

```{r ml-eval-helpers}
#| message: false
#| warning: false

fig_dir <- file.path(res_dir, "figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

# Shared evaluation helpers (level order fixed c("real","fake"); "fake" positive).
extract_binary_metrics <- function(cm, model_name) {
  tibble(
    Model               = model_name,
    Accuracy            = as.numeric(cm$overall["Accuracy"]),
    Precision           = as.numeric(cm$byClass["Precision"]),
    Recall              = as.numeric(cm$byClass["Recall"]),
    F1                  = as.numeric(cm$byClass["F1"]),
    Specificity         = as.numeric(cm$byClass["Specificity"]),
    `Balanced Accuracy` = as.numeric(cm$byClass["Balanced Accuracy"])
  )
}
evaluate_global <- function(pred, truth, model_name) {
  cm <- caret::confusionMatrix(factor(pred,  c("real", "fake")),
                               factor(truth, c("real", "fake")),
                               positive = "fake", mode = "everything")
  extract_binary_metrics(cm, model_name)
}
subclass_breakdown <- function(truth, pred, prob_fake, subclass) {
  tibble(subclass = subclass, truth = truth, pred = pred, prob_fake = prob_fake) %>%
    group_by(subclass) %>%
    summarise(n = n(), Accuracy = mean(pred == truth),
              Predicted_Fake_Rate = mean(pred == "fake"),
              Mean_Prob_Fake = mean(prob_fake), .groups = "drop")
}
acc_bar <- function(df, title) ggplot(df, aes(subclass, Accuracy)) + geom_col() + ylim(0, 1) +
  labs(title = title, x = "Subclass", y = "Accuracy") + theme_minimal()
save_fig <- function(p, name, w = 8, h = 5) {
  ggplot2::ggsave(file.path(fig_dir, paste0(name, ".png")), p, width = w, height = h, dpi = 300)
  ggplot2::ggsave(file.path(fig_dir, paste0(name, ".pdf")), p, width = w, height = h)
}
read_ml_preds <- function(f) readr::read_csv(file.path(art_dir, f), show_col_types = FALSE) %>%
  mutate(truth = factor(truth, c("real", "fake")), pred = factor(pred, c("real", "fake")),
         subclass = factor(subclass, c("HF", "MF", "HR", "MR")))
```

```{r ml-eval}
#| message: true
#| warning: false

ml_spec <- tibble::tribble(
  ~file,                    ~Model,                              ~slug,       ~fig,
  "ml_preds_logit.csv",     "Logistic Regression",               "logit",     "ml_logistic_subclass_accuracy",
  "ml_preds_nb_tfidf.csv",  "Naive Bayes (Gaussian, TF-IDF)",    "nb",        "ml_nb_tfidf_subclass_accuracy",
  "ml_preds_nb_counts.csv", "Naive Bayes (multinomial, counts)", "nb_counts", "ml_nb_counts_subclass_accuracy",
  "ml_preds_rf.csv",        "Random Forest",                     "rf",        "ml_rf_subclass_accuracy"
)

ml_metrics <- list(); ml_subclass <- list()
for (i in seq_len(nrow(ml_spec))) {
  f <- ml_spec$file[i]; nm <- ml_spec$Model[i]; slug <- ml_spec$slug[i]
  stopifnot(file.exists(file.path(art_dir, f)))
  d  <- read_ml_preds(f)
  m  <- evaluate_global(d$pred, d$truth, nm)
  sb <- subclass_breakdown(d$truth, d$pred, d$prob_fake, d$subclass) %>% mutate(Model = nm)
  ml_metrics[[nm]] <- m; ml_subclass[[nm]] <- sb
  readr::write_csv(m, file.path(tab_dir, paste0(slug, "_metrics.csv")))
  readr::write_csv(sb %>% select(subclass, n, Accuracy, Predicted_Fake_Rate, Mean_Prob_Fake),
                   file.path(tab_dir, paste0(slug, "_by_subclass.csv")))
  save_fig(acc_bar(sb, paste0(nm, ": Accuracy by Subclass")), ml_spec$fig[i])
}

# Explicit, self-documenting NB names + the two-row NB comparison, written alongside
# the per-model tables (matches the thesis terminology: Gaussian-on-TF-IDF vs multinomial-on-counts).
nb_g <- "Naive Bayes (Gaussian, TF-IDF)"; nb_m <- "Naive Bayes (multinomial, counts)"
readr::write_csv(ml_metrics[[nb_g]],  file.path(tab_dir, "nb_gaussian_tfidf_metrics.csv"))
readr::write_csv(ml_subclass[[nb_g]], file.path(tab_dir, "nb_gaussian_tfidf_by_subclass.csv"))
readr::write_csv(ml_metrics[[nb_m]],  file.path(tab_dir, "nb_multinomial_counts_metrics.csv"))
readr::write_csv(ml_subclass[[nb_m]], file.path(tab_dir, "nb_multinomial_counts_by_subclass.csv"))
readr::write_csv(bind_rows(ml_metrics[[nb_g]], ml_metrics[[nb_m]]),
                 file.path(tab_dir, "nb_compare_metrics.csv"))

# Traditional-ML comparison tables — read back by 03 to fold in BERT.
traditional_metrics <- bind_rows(ml_metrics)
traditional_subclass_results <- bind_rows(ml_subclass) %>%
  select(Model, subclass, n, Accuracy, Predicted_Fake_Rate, Mean_Prob_Fake)
readr::write_csv(traditional_metrics,          file.path(tab_dir, "traditional_metrics.csv"))
readr::write_csv(traditional_subclass_results, file.path(tab_dir, "traditional_subclass_results.csv"))
print(knitr::kable(traditional_metrics, digits = 3))
```

```{r ml-rf-figures}
#| message: true
#| warning: false

# RF internals (model-based; written to ../artifacts above) -> figures.
if (file.exists(file.path(art_dir, "rf_importance.csv"))) {
  rf_importance <- readr::read_csv(file.path(art_dir, "rf_importance.csv"), show_col_types = FALSE)
  readr::write_csv(rf_importance, file.path(tab_dir, "rf_importance.csv"))
  save_fig(ggplot(head(rf_importance, 20),
                  aes(reorder(Feature, MeanDecreaseAccuracy), MeanDecreaseAccuracy)) +
             geom_col() + coord_flip() +
             labs(title = "Random Forest: Top-20 features (permutation importance)",
                  x = NULL, y = "Mean decrease in accuracy") + theme_minimal(),
           "ml_rf_importance_top20", w = 8, h = 6)
}
if (file.exists(file.path(art_dir, "rf_oob.csv"))) {
  rf_oob <- readr::read_csv(file.path(art_dir, "rf_oob.csv"), show_col_types = FALSE)
  readr::write_csv(rf_oob, file.path(tab_dir, "rf_oob.csv"))
  save_fig(ggplot(rf_oob, aes(trees, OOB)) + geom_line() +
             labs(title = "Random Forest OOB Error Curve",
                  x = "Number of trees", y = "OOB error") + theme_minimal(),
           "ml_rf_oob_curve")
}
```