---
title: "LLM inference — Claude Haiku 4.5 (Anthropic, Batch API)"
format: html
editor: visual
---

# What this file is

A **standalone runner** for **Claude Haiku 4.5 (Anthropic)** — one model, so it never blocks behind the others. It reads the shared data (`../artifacts/gossip_model_data.csv`) and the frozen exemplar IDs (`../artifacts/llm_fewshot_exemplars_ids.csv`), reconstructs the exemplar text from the train split, and rebuilds the **identical** prompts every model uses, so rows and text are byte-for-byte the same. It writes:

- `../artifacts/llm_preds_claude_haiku_zero.csv`
- `../artifacts/llm_preds_claude_haiku_few.csv`

When both exist, run [`04b_llm_eval.qmd`](04b_llm_eval.qmd) (it picks them up by filename).

> **How to run.** Open in RStudio and **Run All Chunks** (Ctrl+Alt+R); do not `quarto render`. Claude Haiku uses Anthropic's **Message Batches API** (temperature 0): the job runs on Anthropic's servers and this file only submits then polls, saving the batch id to a `.json` state file. It is safe to interrupt — re-running resumes polling from that state (no re-submission, no double charge), and any regime whose output CSV already exists is skipped. Runs independently of the OpenAI and Gemini runners (different files, no collisions).

# Setup: libraries

```{r v192-libs}
#| message: false
#| warning: false

# Standalone session: load everything this file needs on its own.
# IMPORTANT: `withr` must be ATTACHED (not merely installed). ellmer 0.4.1's
# batch code calls local_tempfile() without importing {withr}; attaching it here
# lets that unqualified call resolve via the search path — for BOTH submitting a
# batch AND retrieving a finished one. (This single line is what TFM_V19.1 had
# and the first version of this file was missing.)
pkgs <- c("tidyverse", "ellmer", "withr")
missing <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(missing) > 0) install.packages(missing)
invisible(lapply(pkgs, library, character.only = TRUE))

cat("Loaded:", paste(pkgs, collapse = ", "), "\n")
```

# Shared data (same source of truth as every other model)

```{r v192-data}
#| message: true
#| warning: false

# ── Read the single shared modelling table; rebuild the exact test/train frames ─────────
model_data_path <- "../artifacts/gossip_model_data.csv"
stopifnot(file.exists(model_data_path))

model_data <- readr::read_csv(model_data_path, show_col_types = FALSE) %>%
  mutate(
    subclass = factor(subclass, levels = c("HF", "MF", "HR", "MR")),
    label    = factor(label,    levels = c("fake", "real")),
    split    = factor(split,    levels = c("train", "validation", "test"))
  )

test_df  <- model_data %>% filter(split == "test")    # 2,796 rows — scored by all models
train_df <- model_data %>% filter(split == "train")   # 60% — few-shot exemplars come from here

cat("Loaded", nrow(model_data), "rows  |  test =", nrow(test_df),
    " train =", nrow(train_df), "\n")
```

# The 8 frozen few-shot exemplars (identical to the main doc)

```{r v192-exemplars}
#| message: true
#| warning: false

llm_test <- test_df %>% select(doc_id, subclass, label, doc_text)
stopifnot(nrow(llm_test) == 2796)

# Frozen 8 few-shot exemplars (2 per subclass). Only doc_id+subclass+label are
# committed (../artifacts/llm_fewshot_exemplars_ids.csv) — no article text on
# GitHub — and the text is reconstructed by joining those IDs to train_df. If the
# IDs file is absent, the identical 8 are regenerated from the fixed seed.
exemplar_ids_path <- "../artifacts/llm_fewshot_exemplars_ids.csv"
if (file.exists(exemplar_ids_path)) {
  exemplar_ids <- readr::read_csv(exemplar_ids_path, show_col_types = FALSE)
} else {
  set.seed(2026)
  exemplar_ids <- train_df %>%
    group_by(subclass) %>%
    slice_sample(n = 2) %>%
    mutate(rep = row_number()) %>%
    ungroup() %>%
    arrange(rep, match(subclass, c("HF", "HR", "MF", "MR"))) %>%
    transmute(doc_id, subclass = as.character(subclass), label = as.character(label))
  readr::write_csv(exemplar_ids, exemplar_ids_path)
}

# Recover exemplar text from train_df by doc_id, preserving the frozen order.
exemplars <- exemplar_ids %>%
  mutate(subclass = factor(subclass, levels = c("HF", "MF", "HR", "MR")),
         label    = as.character(label)) %>%
  left_join(train_df %>% select(doc_id, doc_text), by = "doc_id")
stopifnot(!any(is.na(exemplars$doc_text)))   # every exemplar id must be in train_df

stopifnot(nrow(exemplars) == 8)
stopifnot(length(intersect(exemplars$doc_id, llm_test$doc_id)) == 0)   # no leakage
cat("Exemplar order (label):", paste(exemplars$label, collapse = ", "), "\n")
```

# Prompts, model registry, keys

```{r v192-prompts}
#| message: true
#| warning: false

sys_base <- paste(
  "Read the following news article and determine whether it is real or fake",
  "Answer with exactly one word, lowercase, with no punctuation and no explanation:",
  "either fake or real.",
  sep = "\n"
)

render_exemplars <- function(ex) {
  blocks <- purrr::pmap_chr(
    ex[c("doc_text", "label")],
    function(doc_text, label) paste0("Article:\n", doc_text, "\n\nAnswer: ", label)
  )
  paste0(
    "Here are labelled examples:\n\n",
    paste(blocks, collapse = "\n\n---\n\n"),
    "\n\nNow classify the next article in the same way."
  )
}

sys_zero <- sys_base
sys_few  <- paste(sys_base, render_exemplars(exemplars), sep = "\n\n")

test_prompts <- as.list(paste0("Article:\n", llm_test$doc_text))
stopifnot(length(test_prompts) == nrow(llm_test))

models_spec <- tibble::tribble(
  ~key,           ~provider,    ~model,                      ~method,
  "gpt4o_mini",   "openai",     "gpt-4o-mini-2024-07-18",     "batch",
  "claude_haiku", "anthropic",  "claude-haiku-4-5-20251001",  "batch",
  "gemini_flash", "gemini",     "gemini-2.5-flash",           "parallel"
)

load_api_keys <- function(path = "API Keys.txt") {
  if (!file.exists(path)) {
    stop("API key file not found: ", normalizePath(path, mustWork = FALSE),
         "\n  -> create 'API Keys.txt' with lines OPENAI_API_KEY=, ANTHROPIC_API_KEY=, GEMINI_API_KEY=")
  }
  lines <- trimws(readLines(path, warn = FALSE))
  lines <- lines[nzchar(lines) & !startsWith(lines, "#") & grepl("=", lines, fixed = TRUE)]
  env <- setNames(trimws(sub("^[^=]*=", "", lines)),
                  trimws(sub("=.*$", "", lines)))
  keys <- list(
    openai    = unname(env["OPENAI_API_KEY"]),
    anthropic = unname(env["ANTHROPIC_API_KEY"]),
    gemini    = unname(env["GEMINI_API_KEY"])
  )
  miss <- names(keys)[vapply(keys, function(k) is.na(k) || !nzchar(k), logical(1))]
  if (length(miss) > 0) {
    stop("API Keys.txt is missing/blank for: ", paste(miss, collapse = ", "),
         " (expected OPENAI_API_KEY / ANTHROPIC_API_KEY / GEMINI_API_KEY)")
  }
  keys
}

make_chat <- function(provider, model, system_prompt, keys) {
  switch(provider,
    openai    = chat_openai(model = model, system_prompt = system_prompt,
                            credentials = function() keys$openai,    params = params(temperature = 0)),
    anthropic = chat_anthropic(model = model, system_prompt = system_prompt,
                               credentials = function() keys$anthropic, params = params(temperature = 0)),
    gemini    = chat_google_gemini(model = model, system_prompt = system_prompt,
                                   credentials = function() keys$gemini, params = params(temperature = 0)),
    stop("unknown provider: ", provider)
  )
}

write_preds <- function(key, setting_name, raw) {
  out_csv <- sprintf("../artifacts/llm_preds_%s_%s.csv", key, setting_name)
  stopifnot(length(raw) == nrow(llm_test))             # positional alignment
  tibble(
    doc_id   = llm_test$doc_id,
    subclass = as.character(llm_test$subclass),
    truth    = as.character(llm_test$label),
    setting  = setting_name,
    model    = key,
    raw      = raw
  ) %>% readr::write_csv(out_csv)
  message("✓ wrote ", out_csv, "  (", length(raw), " rows)")
}
```

# Run Claude Haiku 4.5 (zero-shot and 8-shot) — resumable

```{r v192-run}
#| message: true
#| warning: false

# ── Belt-and-suspenders for the ellmer 0.4.1 local_tempfile bug ───────────────
# Attaching withr (libraries chunk) is the real fix and covers every code path.
# We ALSO shim withr::local_tempfile straight into the Anthropic batch_submit
# method, in case withr is ever not on the search path. Wrapped so a name change
# in ellmer internals can never stop the run.
stopifnot(requireNamespace("withr", quietly = TRUE))
local({
  tryCatch({
    gen <- ellmer:::batch_submit
    cls <- ellmer:::ProviderAnthropic
    m   <- S7::method(gen, cls)
    env <- environment(m)
    if (is.null(env) || !exists("local_tempfile", envir = env, inherits = FALSE)) {
      shim <- new.env(parent = env)
      shim$local_tempfile <- withr::local_tempfile
      environment(m) <- shim
      S7::method(gen, cls) <- m
      message("[patch] shimmed withr::local_tempfile into Anthropic batch_submit.")
    }
  }, error = function(e)
    message("[patch] Anthropic shim skipped (", conditionMessage(e),
            "); relying on attached withr."))
})

api_keys <- load_api_keys()                  # errors early if the key file is missing
this_key <- "claude_haiku"
spec     <- models_spec %>% filter(key == this_key)
settings <- list(zero = sys_zero, few = sys_few)

# Submit (or resume) one Batch job, poll until done, write the CSV.
# Resumes automatically on re-run: the .json state file holds the batch id.
run_batch_setting <- function(setting_name, wait_s = 60) {
  out_csv <- sprintf("../artifacts/llm_preds_%s_%s.csv", this_key, setting_name)
  if (file.exists(out_csv)) {
    message("✓ already complete, skipping: ", out_csv); return(invisible())
  }

  chat       <- make_chat(spec$provider, spec$model, settings[[setting_name]], api_keys)
  state_path <- sprintf(".llm_%s_%s.json", this_key, setting_name)
  message(">> ", this_key, " / ", setting_name,
          "  (batch state: ", state_path, ")  submitting / resuming…")

  repeat {
    done <- tryCatch({
      raw <- batch_chat_text(chat, test_prompts, path = state_path)
      write_preds(this_key, setting_name, raw)
      TRUE
    }, error = function(e) {
      msg <- conditionMessage(e)
      # Only swallow transient NETWORK errors; auth / hash-mismatch are real.
      if (grepl("resolve host|Failed to perform HTTP|Timeout|timed out|curl|Could not resolve|SSL|connection",
                msg, ignore.case = TRUE)) {
        message(sprintf("[net] %s\n      retrying in %ds (the batch keeps running on Anthropic's servers)…",
                        msg, wait_s))
        Sys.sleep(wait_s); FALSE
      } else stop(e)
    })
    if (isTRUE(done)) break
  }
}

for (s in names(settings)) run_batch_setting(s)

message("\nClaude Haiku 4.5 done. Outputs: ",
        paste(sprintf("llm_preds_%s_%s.csv", this_key, names(settings)), collapse = ", "),
        "\nNext: run 04a_llm_infer_openai.qmd (GPT-4o-mini) and 04a_llm_infer_gemini.qmd (Gemini), then analysis/04b_llm_eval.qmd.")
```
