---
title: "04b_llm_eval"
format: html
editor: visual
---

The LLM runners (`04a_llm_infer_*.qmd`) write raw one-word answers to `../artifacts/`. Here we parse them to labels — scoring any unresolved row as incorrect, so a parse failure can never inflate accuracy — and evaluate with the same helpers as `02` (ML) and `03` (BERT). No probability is available, so the per-subclass `Predicted_Fake_Rate` carries the detector-bias signal and `Mean_Prob_Fake` is not interpreted for the LLMs. This notebook needs only the committed prediction files — no dataset, no API access. The final chunk reads the all-detector table `03` wrote and folds in the LLM rows.

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(caret)

art_dir <- "../artifacts"
res_dir <- "../results"
tab_dir <- file.path(res_dir, "tables")
fig_dir <- file.path(res_dir, "figures")
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

# ── Shared evaluation helpers  ─────────────────────────────────
extract_binary_metrics <- function(cm, model_name) {
  tibble(
    Model               = model_name,
    Accuracy            = as.numeric(cm$overall["Accuracy"]),
    Precision           = as.numeric(cm$byClass["Precision"]),
    Recall              = as.numeric(cm$byClass["Recall"]),
    F1                  = as.numeric(cm$byClass["F1"]),
    Specificity         = as.numeric(cm$byClass["Specificity"]),
    `Balanced Accuracy` = as.numeric(cm$byClass["Balanced Accuracy"])
  )
}
evaluate_global <- function(pred, truth, model_name) {
  cm <- caret::confusionMatrix(factor(pred,  c("real", "fake")),
                               factor(truth, c("real", "fake")),
                               positive = "fake", mode = "everything")
  extract_binary_metrics(cm, model_name)
}
subclass_breakdown <- function(truth, pred, prob_fake, subclass) {
  tibble(subclass = subclass, truth = truth, pred = pred, prob_fake = prob_fake) %>%
    group_by(subclass) %>%
    summarise(n = n(), Accuracy = mean(pred == truth),
              Predicted_Fake_Rate = mean(pred == "fake"),
              Mean_Prob_Fake = mean(prob_fake), .groups = "drop")
}

# ── Model registry (keys to find files; pretty labels for tables) ─────────────
llm_keys     <- c("gpt4o_mini", "claude_haiku", "gemini_flash")
llm_settings <- c("zero", "few")
model_labels <- c(gpt4o_mini = "GPT-4o-mini", claude_haiku = "Claude Haiku 4.5",
                  gemini_flash = "Gemini 2.5 Flash")
pretty_name <- function(key, setting)
  paste0(model_labels[[key]], " (", c(zero = "zero-shot", few = "8-shot")[[setting]], ")")

# Strict parser: exactly one of "fake"/"real" -> that label; both/neither -> NA.
parse_label <- function(x) {
  t <- str_to_lower(str_trim(ifelse(is.na(x), "", x)))
  dplyr::case_when(str_detect(t, "fake") & !str_detect(t, "real") ~ "fake",
                   str_detect(t, "real") & !str_detect(t, "fake") ~ "real",
                   TRUE ~ NA_character_)
}

# Human-adjudicated review of parse failures (auditable, optional, preserves edits).
review_path <- file.path(art_dir, "llm_parse_review.csv")
load_overrides <- function(path = review_path) {
  if (!file.exists(path)) return(NULL)
  ov <- readr::read_csv(path, show_col_types = FALSE) %>%
    filter(!is.na(manual_label) & nzchar(trimws(manual_label))) %>%
    transmute(model, setting, doc_id, manual_label = str_to_lower(trimws(manual_label)))
  if (nrow(ov) == 0) return(NULL)
  bad <- setdiff(unique(ov$manual_label), c("fake", "real"))
  if (length(bad) > 0) stop("manual_label must be 'fake'/'real'; got: ", paste(bad, collapse = ", "))
  ov
}

# Final label = parser result, else manual label; still-unresolved -> scored WRONG.
load_llm_preds <- function(key, setting_name, overrides = load_overrides()) {
  f <- file.path(art_dir, sprintf("llm_preds_%s_%s.csv", key, setting_name))
  if (!file.exists(f)) return(NULL)
  d <- readr::read_csv(f, show_col_types = FALSE) %>%
    mutate(model = key, setting = setting_name, auto_parsed = parse_label(raw))
  ov <- if (is.null(overrides)) NULL else filter(overrides, model == key, setting == setting_name)
  if (!is.null(ov) && nrow(ov) > 0) d <- left_join(d, ov[c("doc_id", "manual_label")], by = "doc_id")
  else d$manual_label <- NA_character_
  d %>% mutate(
    final    = dplyr::coalesce(auto_parsed, manual_label),
    auto_ok  = !is.na(auto_parsed),
    resolved = !is.na(final),
    truth    = factor(truth, levels = c("real", "fake")),
    pred_chr = dplyr::if_else(resolved, final, dplyr::if_else(truth == "fake", "real", "fake")),
    pred     = factor(pred_chr, levels = c("real", "fake")),
    subclass = factor(subclass, levels = c("HF", "MF", "HR", "MR"))
  )
}
```

## Global + subclass metrics

```{r llm-eval}
#| message: true
#| warning: false

llm_pred_list    <- list()
llm_metrics_list <- list()
for (k in llm_keys) for (s in llm_settings) {
  d <- load_llm_preds(k, s)
  if (is.null(d)) { message("· missing LLM predictions: ", k, " ", s); next }
  nm <- pretty_name(k, s)
  llm_pred_list[[nm]] <- d
  cat(sprintf("%-26s auto-parse fail: %.2f%% | unresolved after review: %d\n",
              nm, 100 * mean(!d$auto_ok), sum(!d$resolved)))
  llm_metrics_list[[nm]] <- evaluate_global(d$pred, d$truth, nm)
}

if (length(llm_metrics_list) > 0) {
  llm_metrics <- bind_rows(llm_metrics_list)
  readr::write_csv(llm_metrics, file.path(tab_dir, "llm_metrics.csv"))

  llm_subclass_results <- purrr::imap_dfr(llm_pred_list, function(d, nm)
    subclass_breakdown(d$truth, d$pred, as.numeric(d$pred == "fake"), d$subclass) %>%
      mutate(Model = nm)) %>%
    select(Model, subclass, n, Accuracy, Predicted_Fake_Rate)
  readr::write_csv(llm_subclass_results, file.path(tab_dir, "llm_subclass_results.csv"))

  # Parse-reliability summary (one row per model x regime)
  purrr::imap_dfr(llm_pred_list, function(d, nm) tibble(
    Model = nm, auto_parse_fail_pct = round(100 * mean(!d$auto_ok), 3),
    unresolved_after_review = sum(!d$resolved))) %>%
    readr::write_csv(file.path(tab_dir, "llm_parse_reliability.csv"))

  print(knitr::kable(llm_metrics, digits = 3))

  ggplot2::ggsave(file.path(fig_dir, "plot_llm_accuracy_by_subclass.png"),
    ggplot(llm_subclass_results, aes(subclass, Accuracy, fill = Model)) +
      geom_col(position = position_dodge(width = 0.8), width = 0.75) + ylim(0, 1) +
      labs(title = "LLM accuracy by subclass (zero-shot vs 8-shot)",
           x = "Subclass", y = "Accuracy") + theme_minimal(),
    width = 9, height = 5.5, dpi = 300)
  ggplot2::ggsave(file.path(fig_dir, "plot_llm_predicted_fake_rate.png"),
    ggplot(llm_subclass_results, aes(subclass, Predicted_Fake_Rate, fill = Model)) +
      geom_col(position = position_dodge(width = 0.8), width = 0.75) +
      geom_hline(yintercept = 0.5, linetype = "dashed") + ylim(0, 1) +
      labs(title = "LLM predicted-fake rate by subclass (detector bias)",
           x = "Subclass", y = "Predicted fake rate") + theme_minimal(),
    width = 9, height = 5.5, dpi = 300)

  # Per-model confusion matrices (counts)
  for (nm in names(llm_pred_list)) {
    cm <- caret::confusionMatrix(llm_pred_list[[nm]]$pred, llm_pred_list[[nm]]$truth,
                                 positive = "fake", mode = "everything")
    safe <- gsub("[^A-Za-z0-9]+", "_", nm)
    as.data.frame.matrix(cm$table) %>% tibble::rownames_to_column("prediction\\reference") %>%
      readr::write_csv(file.path(tab_dir, paste0("confusion_", safe, ".csv")))
  }
} else {
  message("[04] No LLM predictions found in ", art_dir, " — run the 04a_llm_infer_*.qmd files first.")
}
```

## Does the 8-shot set help? McNemar per model

```{r llm-mcnemar}
#| message: true
#| warning: false

mcnemar_rows <- list()
for (k in llm_keys) {
  dz   <- llm_pred_list[[pretty_name(k, "zero")]]
  dfew <- llm_pred_list[[pretty_name(k, "few")]]
  if (is.null(dz) || is.null(dfew)) next
  paired <- inner_join(dz %>% transmute(doc_id, truth, zero = pred),
                       dfew %>% transmute(doc_id, few = pred), by = "doc_id") %>%
    mutate(zero_correct = zero == truth, few_correct = few == truth)
  mt <- stats::mcnemar.test(table(paired$zero_correct, paired$few_correct))
  cat(sprintf("%-18s zero=%.3f  8-shot=%.3f  McNemar p=%.4g\n",
              model_labels[[k]], mean(paired$zero_correct), mean(paired$few_correct), mt$p.value))
  mcnemar_rows[[k]] <- tibble(Model = model_labels[[k]],
    Zero_Acc = mean(paired$zero_correct), FewShot_Acc = mean(paired$few_correct),
    Delta = mean(paired$few_correct) - mean(paired$zero_correct), McNemar_p = mt$p.value)
}
if (length(mcnemar_rows) > 0) {
  llm_fewshot_lift <- bind_rows(mcnemar_rows)
  readr::write_csv(llm_fewshot_lift, file.path(tab_dir, "llm_fewshot_lift.csv"))
  print(knitr::kable(llm_fewshot_lift, digits = 4))
}
```

## Full comparison across every detector

```{r full-comparison}
#| message: true
#| warning: false

# Read the all-detector table 03 wrote (traditional ML + BERT) and append the LLMs.
amm <- file.path(tab_dir, "all_model_metrics.csv")
asr <- file.path(tab_dir, "all_subclass_results.csv")
base_metrics  <- if (file.exists(amm)) readr::read_csv(amm, show_col_types = FALSE) else NULL
base_subclass <- if (file.exists(asr)) readr::read_csv(asr, show_col_types = FALSE) else NULL
if (is.null(base_metrics))
  message("[04] all_model_metrics.csv not found — run 02 then 03 first for the full comparison.")

full_metrics <- bind_rows(base_metrics, if (exists("llm_metrics")) llm_metrics)
readr::write_csv(full_metrics, file.path(tab_dir, "full_metrics.csv"))
print(knitr::kable(full_metrics, digits = 3))

full_subclass <- bind_rows(
  if (!is.null(base_subclass)) base_subclass %>% select(Model, subclass, n, Accuracy),
  if (exists("llm_subclass_results")) llm_subclass_results %>% select(Model, subclass, n, Accuracy)
)
readr::write_csv(full_subclass, file.path(tab_dir, "full_subclass.csv"))

ggplot2::ggsave(file.path(fig_dir, "plot_all_detectors_subclass_accuracy.png"),
  ggplot(full_subclass, aes(subclass, Accuracy, fill = Model)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.75) + ylim(0, 1) +
    labs(title = "Subclass accuracy by detector", x = "Subclass", y = "Accuracy") +
    theme_minimal(),
  width = 11, height = 5.5, dpi = 300)
```
